Exploring mathematical objects from custom-tailored mathematical universes

### 2022-04-02 paper-filmat.tex
arXiv version

%\documentclass[oneside,reqno]{amsart}
\documentclass[graybox]{svmult}

\usepackage{type1cm}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage[bottom]{footmisc}
\usepackage{newtxtext}
\usepackage{newtxmath}

@@

@@

\title*{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\institute{Ingo Blechschmidt \at Universität Augsburg, Institut für Mathematik,
Universitätsstr. 14,
86159 Augsburg, Germany, \email{ingo.blechschmidt@math.uni-augsburg.de}}

@@

\abstract{
  Toposes can be pictured as mathematical universes. Besides the standard topos,
  in which most of mathematics unfolds, there is a colorful host of alternate
  toposes in which mathematics plays out slightly
  differently.
  For instance, there are toposes in which the axiom of choice and the
  intermediate value theorem from undergraduate calculus fail.
  The purpose of this contribution is to give a glimpse of the toposophic
  landscape, presenting several specific toposes and exploring their peculiar
  properties, and to explicate how toposes provide distinct lenses through
  which the usual mathematical objects of the standard topos can be viewed.
  \emph{Key words:} topos theory, realism debate, well-adapted language,
  constructive mathematics
}
\medskip

@@

% \begin{document}

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and that we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{The research
program put forward by~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories. \textbf{\textcolor{red}{NOTE TO EDITOR: Insert cross reference to Olivia's chapter of
this book.}}}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

\bigskip
\begin{center}
\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  % FUTURE: Countable axiom of choice
  \bottomrule
\end{tabular}
\end{center}
\bigskip

>>>>

\documentclass[graybox]{svmult}
\pdfoutput=1

\usepackage{amsmath,amssymb,etex,ifxetex}
\usepackage{type1cm}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage[bottom]{footmisc}
\usepackage{newtx}

@@

\usepackage[{colorlinks,citecolor=[rgb]{0,0.1,0}}]{hyperref}

%\usepackage{geometry}
%\geometry{lmargin=4.6cm,rmargin=4.5cm}

@@

\addtocounter{chapter}{3}

\title{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\institute{Ingo Blechschmidt \at Universität Augsburg, Institut für Mathematik,
Universitätsstr. 14,
86159 Augsburg, Germany, \email{ingo.blechschmidt@math.uni-augsburg.de}}

@@

\abstract{
  Toposes can be pictured as mathematical universes. Besides the standard topos,
  in which most of mathematics unfolds, there is a colorful host of alternate
  toposes in which mathematics plays out slightly
  differently.
  For instance, there are toposes in which the axiom of choice and the
  intermediate value theorem from undergraduate calculus fail.
  The purpose of this contribution is to give a glimpse of the toposophic
  landscape, presenting several specific toposes and exploring their peculiar
  properties, and to explicate how toposes provide distinct lenses through
  which the usual mathematical objects of the standard topos can be viewed.
}
\keywords{topos theory, realism debate, well-adapted language,
  constructive mathematics}
\medskip

@@

%\newcommand{\RR}{\mathbb{R}}

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and that we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{The research
program put forward by~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories. \textit{\textcolor{red}{Note to editor: please insert cross reference to Olivia's chapter of
this book.}}}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

\bigskip
\begin{center}
\begin{tabular}{ll@{\qquad}l}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  % FUTURE: Countable axiom of choice
  \bottomrule
\end{tabular}
\end{center}
\bigskip
